---
title: "Efficient loops"
author:   
  - name: Juan R Gonzalez
    affiliation: 
    - Bioinformatics Research Group in Epidemiology, Barcelona Institute for Global Health, Spain
    email: juanr.gonzalez@isglobal.org
output:
  BiocStyle::html_document:
    toc_depth: 2
  BiocStyle::pdf_document:
    toc_depth: 2
---


```{r style, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
library("knitr")
options(width=120)
opts_chunk$set(message = FALSE, error = TRUE, warning = TRUE,
               comment = "")
```


# Intro

There are several packages to perform parallel computin. The reason for using `doParallel` package (and not the parallel package) is that the older parallel package, parallelization was not working entirely on Windows and you had to write different code for it to work. The doParallel package is trying to make it happen on all platforms: UNIX, LINUX and WINDOWS, so it’s a pretty decent wrapper.

# Example

Let us assume we want to compute the first 

```{r function_primers}
getPrimeNumbers <- function(n) {  
   n <- as.integer(n)
   if(n > 1e6) stop("n too large")
   primes <- rep(TRUE, n)
   primes[1] <- FALSE
   last.prime <- 2L
   for(i in last.prime:floor(sqrt(n)))
   {
      primes[seq.int(2L*last.prime, n, last.prime)] <- FALSE
      last.prime <- last.prime + min(which(primes[(last.prime+1):n]))
   }
   which(primes)
}
```

The function is taken from: http://stackoverflow.com/questions/3789968/generate-a-list-of-primes-in-r-up-to-a-certain-number. Now let’s see the code for each type of loops and finally compare the runtime performance in each loop type and how they perform.

# for

```{r for}
forFunction <- function(x) {
  result <- list()  
  for (i in seq_along(x)) {  
    result[[i]] <- getPrimeNumbers(x[i])
  }
  result
}
```

# lapply

```{r lapply}
lapplyFunction <- function(x) {
  result <- lapply(x, getPrimeNumbers)
  result
}
```

# doParallel::parLapply

```{r doParllel_parLapply}
library(doParallel)
ncores <- detectCores() - 1  
registerDoParallel(cores=ncores)  
cl <- makeCluster(ncores)  

parLapplyFunction <- function(x){
  result <- parLapply(cl, x, getPrimeNumbers)  
  result
}
stopCluster(cl)
```  


# doParallel::foreach

```{r foreach}
cl <- makeCluster(ncores)  
registerDoParallel(cl)  
foreachFunction <- function(x) {
  result <- foreach(i=x, .export="getPrimeNumbers") %dopar%
    getPrimeNumbers(i)
  result
}  
stopCluster(cl)
```

# doParallel::mclapply

```{r mclapply}
# Only works in Linux (Windows ncores must be set equal to 1)
result <- mclapply(x, getPrimeNumbers, mc.cores=ncores)
```


# Evaluation

```{r system}
x <- 10:2000
system.time(result <- lapply(x, getPrimeNumbers))
```


```{r microbenchmarck}
library(microbenchmark)
library(doParallel)
ncores <- detectCores() - 1  
registerDoParallel(cores=ncores)  
# cl <- makeCluster(ncores, type="FORK")
cl <- makeCluster(ncores)
registerDoParallel(cl)  

x <- 10:1000
bench <- microbenchmark(forF = forFunction(x),
                        applyF = lapplyFunction(x),
                        forEachF = foreachFunction(x),
                        parLapplyF = parLapplyFunction(x),
                        times=20)
bench
stopCluster(cl)
```

```{r plot}
plot(bench)
```


# Session info

```{r}
sessionInfo()
```




